---
title: "HarmonizeR function, v.0"
author: "C.A. Kiahtipes"
date: "6/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

#HarmonizeR function

harmonizer=function(rec_names,master_file="AML_base_2019.csv",taxa_file=FALSE,Tilia_format=TRUE,APD_format=TRUE){ #Function needs names of pollen records and a master list. ###consider repairing "rec_names" to something more sensible.
  path=getwd()
  ###UNIVERSAL OBJECTS: MASTER POLLEN/MORPHOTYPE LIST
  
  master=read.csv(paste0(master_file),header=TRUE) ###CAK_edits #master file: need systematic naming convention and use across document.#Update to read current list from AML
  if(APD_format==TRUE){
    original_names=unique(master$ORIGINAL.NAMES) ###CAK_edits #master file: need systematic naming convention and use across document.
    proposed_names=unique(master$PROPOSED.NAMES) ###CAK_edits #master file: need systematic naming convention and use across document.
  } else {
    proposed_names=unique(master$PROPOSED.NAMES) ###CAK_edits #master file: need systematic naming convention and use across document. #Other master documents NEED to have this title!
  }
  
  ###COUNT TAXA AND PREPARE TABLES
  
  name_length=vector(mode="numeric",length=1) #vector to track number of short names per record
  fullname_length=vector(mode="numeric",length=1) #vector to track number of long names per record 
  
  for(i in 1:length(rec_names)){ #Read data and get taxa names from columns.
    if(Tilia_format==TRUE){
      pol_data=read.csv(paste0(path,'/data/',rec_names[i],".csv"),header=TRUE,row.names="X")
      pol_data=t(pol_data)
      pol_data=as.data.frame(pol_data)
    } else {
      pol_data=read.csv(paste0(path,'/data/',rec_names[i],"_data.csv"),header=TRUE,row.names="X")
    }
    cn=colnames(pol_data)
    name_length[i]=length(cn)
    if(taxa_file==TRUE){
      pol_names=read.csv(paste0(path,'/data/',rec_names[i],"_taxa.csv"),header=TRUE) 
      fullnames=pol_names$FullNames
      fullname_length[i]=length(fullnames)
      if(name_length[i]-fullname_length[i]!=0){
        print(paste0("Check names in ",rec_names[i]))
      } else {
        print(paste0("Taxa and Data files match, ",rec_names[i]))
      }
    }
  } #Close for loop to track names per record.
  
  #USE COUNTS TO CREATE MATRIX FOR PLACING DATA
  
  #Use name_length to build out a matrix.
  taxa_just=matrix(ncol=5,nrow=sum(name_length)) 
  taxa_just=as.data.frame(taxa_just)
  colnames(taxa_just)=c("REC_no","CODE_APD","SHORT_TAXON","FULL_TAXON","COMMENT") 

  
  for(i in 1:length(rec_names)){ #For loop reads each record and writes taxa names into it. 
    print(rec_names[i])
    if(Tilia_format==TRUE){
      pol_data=read.csv(paste0(path,'/data/',rec_names[i],".csv"),header=TRUE,row.names="X")
      pol_data=t(pol_data)
      pol_data=as.data.frame(pol_data)
    } else {
      pol_data=read.csv(paste0(path,'/data/',rec_names[i],"_data.csv"),header=TRUE,row.names="X")
    }
    cn=colnames(pol_data)
    
    if(taxa_file==TRUE){
      taxa_full=read.csv(paste0(path,'/data/',rec_names[i],"_taxa.csv"),header=TRUE)
      fullnames=taxa_full$FullNames
    } else {}
    
    if(i==1){ #Adding numbers to track records
      begin=1
      end=name_length[i]
      taxa_just[begin:end,1]=rep(i,name_length[i])
    } else {
      begin=sum(name_length[1:(i-1)])+1
      end=sum(name_length[1:i])
      taxa_just[begin:end,1]=rep(i,name_length[i])
    }
    
    if(i==1){ #Adding record codes to list
      begin=1
      end=name_length[i]
      taxa_just[begin:end,2]=rep(rec_names[i],name_length[i]) ###Using APD Codes to name records!!!
    } else {
      begin=sum(name_length[1:(i-1)])+1
      end=sum(name_length[1:i])
      taxa_just[begin:end,2]=rep(rec_names[i],name_length[i])
    }
    ###Need to update this to get flexibility in data formats - Tilia vs. Rioja exports...
    if(taxa_file==TRUE){ #If we have a _taxa.csv file, write short names from the CodeNames entries
      if(i==1){ #Adding shortened names to matrix
        begin=1
        end=name_length[i]
        taxa_just[begin:end,3]=c(taxa_full$CodeNames)
      } else {
        begin=sum(name_length[1:(i-1)])+1
        end=sum(name_length[1:i])
        taxa_just[begin:end,3]=c(taxa_full$CodeNames)
      }
    } else { #If there's no _taxa.csv file, then we don't do anything now and derive short names later...
    }
    
    if(taxa_file==TRUE){ #We write the taxa_just matrix differently depending on what's available... ###CAK 7.6.2021 If we have the taxa file we write it from fullnames.
      if(i==1){ #Adding full names to matrix
        begin=1
        end=name_length[i]
        taxa_just[begin:end,4]=c(fullnames)
      } else {
        begin=sum(name_length[1:(i-1)])+1
        end=sum(name_length[1:i])
        taxa_just[begin:end,4]=c(fullnames)
      }
    } else { #IF there's no _taxa.csv file, treat the column names as the long names.
      if(i==1){ #Adding names to matrix from column names
        begin=1
        end=name_length[i]
        taxa_just[begin:end,4]=c(cn)
      } else {
        begin=sum(name_length[1:(i-1)])+1
        end=sum(name_length[1:i])
        taxa_just[begin:end,4]=c(cn)
      }
    }
    
    
  } #End for loop pulling names and writing into combined table.
  
  ###NEED PROCESS TO GET SHORT NAMES BASED ON SPECIAL CHARACTERS ##CAK 7.6.2021
  
  if(taxa_file==FALSE){ #If there's no taxa file, go ahead and make short names with the long ones available
    short_names=sub('\\ .*|\\_.*|\\/.*|\\-.*','',taxa_just$FULL_TAXON)
    taxa_just[,3]=short_names
  } else { #If there is a taxa file, go ahead and clean up the short names available
    taxa_just$SHORT_TAXON=sub('\\ .*|\\_.*|\\/.*|\\-.*','',taxa_just$SHORT_TAXON)
  }

  ###MATCHING LISTED TAXA WITH MASTER LIST: SETTING UP MATCHING MATRICES
  
  n_matches=vector("numeric",length=length(taxa_just$ORIG_TAXON)) #tracking how many matches there are for each taxon, will pull the max value to size the matrix into which results are written.
  
  for(i in 1:nrow(taxa_just)){ #Defining size of matrix by maximum length of matches.
    if(APD_format==TRUE){
      y=grep(taxa_just$SHORT_TAXON[i],proposed_names) #Use grep to search taxon names in AML's list
      z=grep(taxa_just$SHORT_TAXON[i],original_names) #Use grep to search taxon names in AML's list
      
      if(length(y)>length(z)){ #I want to take the larger of the two lists of matches...
        n_matches[i]=length(y) #...and write it into the n_matches vector.
      } else {
        n_matches[i]=length(z)
      }
      
    } else {
      y=grep(taxa_just$SHORT_TAXON[i],proposed_names) #Use grep to search taxon names in master list
      n_matches[i]=length(y)
    }
    
  } #end of loop searching by taxon
  
  ###SEARCHING TAXON NAMES AND WRITING MATCH MATRIX
  
  match_matrix=matrix(nrow=length(n_matches),ncol=max(n_matches,na.rm=TRUE))
  match_matrix=as.data.frame(match_matrix)
  
  ###Need to accommodate variation in data submitted - should be able to run on just a list of names from a data table...
  
  if(APD_format==TRUE){ #If we're working with APD format, then we can search both original and proposed names from AML's list.
    for(i in 1:nrow(taxa_just)){ #For loop searching for original taxa names, and matching them to revised names.
      y=grep(taxa_just$FULL_TAXON[i],proposed_names) #Starting with ideal match - full name to revised name in AMLs list. ###NAMING OF MASTER LIST ITEMS FOR PORTABILITY?
      if(length(y)==0){ #If there's no matches, let's try our second-tier match - full name to an original name
        y=grep(taxa_just$FULL_TAXON[i],original_names) ###ONLY POSSIBLE WITH AML'S LIST - LOGICAL CONTROL FOR THIS?
        if(length(y)==0){ #If we come up short on our second-tier match, then we try our third tier - the short names.
          y=grep(taxa_just$SHORT_TAXON[i],proposed_names)
          if(length(y)==0){ #If we come up short on our third-tier match, then we try our fourth tier - short names vs. original taxa
            y=grep(taxa_just$SHORT_TAXON[i],original_names)
            if(length(y)==0){ #If our last match should fail..
              match_matrix[i,]=NA #Write in NA for suggestions
              taxa_just$COMMENT[i]="no match, check spelling" ###Write in suggestion
            } else { #If our fourth tier search succeeds...
              for(j in 1:length(y)){ #For loop to handle multiple matches.
                z=grep(original_names[y[j]],master$ORIGINAL.NAMES)
                match_matrix[i,j]=master$PROPOSED.NAMES[z[j]] #And we want to write each of those matches to the matrix.
              }
              taxa_just$COMMENT[i]="update taxonomy, revise upwards?" ###Write in suggestion
            }
          } else { #If we succeed at our third tier match, then we write it.
            for(j in 1:length(y)){ #For loop to handle multiple matches.
              match_matrix[i,j]=proposed_names[y[j]] #And we want to write each of those matches to the matrix.
            }
            taxa_just$COMMENT[i]="revise upwards?"  ###Write in suggestion
          }
        } else { #If we succeed at second tier, write it
          for(j in 1:length(y)){ #For loop to handle multiple matches.
            z=grep(original_names[y[j]],master$ORIGINAL.NAMES)
            match_matrix[i,j]=master$PROPOSED.NAMES[z[j]] #And we want to write each of those matches to the matrix.
          }
          taxa_just$COMMENT[i]="old taxon, update taxonomy"  ###Write in suggestion
        }
      } else { #If there are matches y will be >0 and we write the results
        #Can we refine this a bit further - exact matches with nomenclatural differences?
        z=grep(paste0('^',taxa_just$FULL_TAXON[i],'$'),proposed_names[y]) #Search for exact match!
        if(length(z)==0){ #if no exact match...
          z=grep(paste0('^',taxa_just$FULL_TAXON[i],'-type','$'),proposed_names[y]) #Try searching for "-type" variations
          if(length(z)==0){ #if appending "-type" doesn't do anything...
            for(j in 1:length(y)){ #For loop to handle multiple matches.
              match_matrix[i,j]=proposed_names[y[j]] #And we want to write each of those matches to the matrix.
            } #End of for loop handling mutliple matches
            taxa_just$COMMENT[i]="close! check nomenclature"  ###Write in suggestion
          } else { #If appending "=type" does work
            match_matrix[i,1]=proposed_names[y[z]]
            taxa_just$COMMENT[i]="match! is -type appropriate?"  ###Write in suggestion
          }
        } else { #If there's an exact match of the full name then write it!
          match_matrix[i,1]=proposed_names[y[z]]
          taxa_just$COMMENT[i]="match! good job"  ###Write in suggestion
        }
      }
    } #End of loop matching original and revised names
    
  } else { #If we're not using the APD format, then searching could be a bit simpler
  
    for(i in 1:nrow(taxa_just)){ #For loop searching for original taxa names, and matching them to revised names.
      y=grep(taxa_just$FULL_TAXON[i],proposed_names) #Starting with ideal match - full name to revised name in AMLs list. ###NAMING OF MASTER LIST ITEMS FOR PORTABILITY?
      if(length(y)==0){ #If there's no matches, let's try our second-tier match - full name to an original name
        y=grep(taxa_just$SHORT_TAXON[i],proposed_names) ###ONLY POSSIBLE WITH AML'S LIST - LOGICAL CONTROL FOR THIS?
        if(length(y)==0){ #If we come up short on our second-tier match, then we try our third tier - the short names.
          match_matrix[i,]=NA #Write in NA for suggestions
          taxa_just$COMMENT[i]="no match, check spelling" ###Write in suggestion
        } else { #If we succeed at second tier, write it
          for(j in 1:length(y)){ #For loop to handle multiple matches.
            z=grep(proposed_names[y[j]],master$ORIGINAL.NAMES)
            match_matrix[i,j]=master$PROPOSED.NAMES[z[j]] #And we want to write each of those matches to the matrix.
          }
          taxa_just$COMMENT[i]="old taxon, update taxonomy"  ###Write in suggestion
        }
      } else { #If there are matches y will be >0 and we write the results
        #Can we refine this a bit further - exact matches with nomenclatural differences?
        z=grep(paste0('^',taxa_just$FULL_TAXON[i],'$'),proposed_names[y]) #Search for exact match!
        if(length(z)==0){ #if no exact match...
          z=grep(paste0('^',taxa_just$FULL_TAXON[i],'-type','$'),proposed_names[y]) #Try searching for "-type" variations
          if(length(z)==0){ #if appending "-type" doesn't do anything...
            for(j in 1:length(y)){ #For loop to handle multiple matches.
              match_matrix[i,j]=proposed_names[y[j]] #And we want to write each of those matches to the matrix.
            } #End of for loop handling mutliple matches
            taxa_just$COMMENT[i]="close! check nomenclature"  ###Write in suggestion
          } else { #If appending "=type" does work
            match_matrix[i,1]=proposed_names[y[z]]
            taxa_just$COMMENT[i]="match! is -type appropriate?"  ###Write in suggestion
          }
        } else { #If there's an exact match of the full name then write it!
          match_matrix[i,1]=proposed_names[y[z]]
          taxa_just$COMMENT[i]="match! good job"  ###Write in suggestion
        }
      }
    } #End of loop matching original and revised names
    
  }
  
  match_colnames=vector(mode="character",length=1)
  
  for(i in 1:(ncol(match_matrix))){ 
    x=paste0("sugg_taxon_",i)
    match_colnames[i]=x
  }
  
  colnames(match_matrix)=match_colnames
  
  output=cbind(taxa_just,match_matrix)
  
  count_NA=table(is.na(output$sugg_taxon_1))
  
  print(paste0(count_NA[2],' unmatched taxa'))
  
  write.csv(output,file="harmonizeR_output.csv",row.names=FALSE)
  
} #Closing out function.

example=read.csv("data/G16867-1.csv",header=TRUE,row.names="X")

```

## HarmonizeR for African Pollen Morphotypes: Tutorial

This is a first draft of an R function I have written to assist with checking African pollen taxon designations against an accepted list of morphotypes.

The goal of this function is to provide a rapid assessment of a set of pollen taxa that are either being published or uploaded to Neotoma. Its operation is relatively simple and it searches the taxon names from the submitted dataset in Anne-Marie Lézine's expansive pollen morphotype list. Her list is especially useful because it contains a column of "ORIGINAL NAMES" and "PROPOSED NAMES". By searching against this column, we can rapidly update older records. This can also be used to rapidly check your own results.

My hope is that this reduces the time that students and researchers spend copy-pasting names or searching spreadsheets and increases the time that we can spend digging into the botanical and taxonomic resources available to us.

### SETUP

Setup is simple. Make sure the harmonizeR.R script and the AML_base_2019.csv files are in your working directory. Whatever datasets you are using need to be saved in a folder called "data".

My preference is to save files in the data folder using their APD code, just like their Tilia counterparts. In this example, we're using "G16867-1". 

### FORMATTING

The harmonizeR script is looking for a .csv file in the data folder containing pollen taxa and counts. This can be formatted with taxa either as rows (as in Tilia) or as columns. By default, the script is looking for a single file.

Regardless of whether the taxa are organized by columns or rows, harmonizeR is expecting the first row of the first column to be blank and designates this as the row names.

```{r original table}

print("G161671-1 original data table")

print(example)
```

### USE

harmonizer(rec_names,master_file,taxa_file=FALSE,Tilia_format=TRUE,APD_format=TRUE)

  rec_names = a single name or concatenated string of names with record names that are to be evaluated.
  
  master_file = a character string identifying the master reference file. Default is "AML_base_2019.csv"
  
  taxa_file = logical indicating whether or not files with taxon information are to be used, specific to CAKs projects. Default is set to FALSE.
  
  Tilia_format = logical indicating whether taxa are organized along the rows (=TRUE) or the columns (=FALSE) of the spreadsheet. Default is set to TRUE.
  
  APD_format = logical indicating whether master spreadsheet contains "ORIGINAL" and "SUGGESTED" columns sensu AML's list. Default is set to TRUE.
  
### TUTORIAL

Here's a test case with a dataset (G16867-1.csv) that I've modified slightly to include a couple common errors (spelling, old taxonomy).

rec_names = "G16867-1"

harmonizer(rec_names)

HarmonizeR combs through the file, searching the taxa names against AML's list and coming up with suggestions. It should give some brief output in the console saying:

"G16867-1"
"2 unmatched taxa"

```{r easy use, echo=TRUE}

harmonizer(rec_names="G16867-1")

```

### EVALUATION AND UPDATING TAXA

Open the "harmonizeR_output.csv" file that's been saved in the working directory. This spreadsheet starts with five columns listing the count of this record in the total records submitted (more useful with multiple records), the APD code of the record (also just the record name), the short taxon name (leading character string), the full taxon name (complete original name in data file), and some commentary from the function on what kind of error you might be looking at. 

```{r table preview, echo=TRUE}

output=read.csv("harmonizeR_output.csv",header=TRUE)

output[1:2,1:5]

```

Following this is a long list of columns with iterative column names (sugg_taxon_1, sugg_taxon_2...) filled with the matches from searching against the master spreadsheet.

At this point, it is up to the user to check the names of their pollen taxa and decide which suggested name is appropriate. It is important to remember that taxonomy is complicated and so are the kind of errors we grapple with. Thus, it is important to know that this function is designed to identify a range of suggestions and while they are typically listed from the most to least relevant, this is not always the case.

### LOCATING ERRORS

HarmonizeR output can be evaluated as a spreadsheet in another program or within R.

```{r no matches, echo=TRUE}

output[is.na(output$sugg_taxon_1),3:6]

```

Above, we can see that we have a spelling error and a name for a row containing chronological information. One will need to be repaired and the other should simply be ignored.

```{r exact match, echo=TRUE}

output[grep("match!",output$COMMENT),3:6]

```

Above, we see some positive news and there's a good number of matches in our list.

```{r partial match, echo=TRUE}

output[output$COMMENT=="revise upwards?",3:15]

```

Above, we're getting matches for the first character string, matching some genera (Celtis & Euphorbia) as well as a family-level designation (Cyperaceae). All of these have been given an essentially meaningless taxonomic addition "(Africa)". We shouldn't follow harmonizeR's suggestion here strictly and can see that there's good suggestions provided.

Note that in this case the second suggestion is superior for Celtis and Euphorbia, while the first suggestion is the best (and only) for Cyperaceae.

```{r taxonomic updates, echo=TRUE}

output[output$COMMENT=="old taxon, update taxonomy",3:15]

```

Above, we have errors where the pollen taxon matched with an older designation and there is a suggested update from AML's list. There's some helpful suggestions for formatting these family-level and subfamily-level designations.




